---
title: "Data Summary"
subtitle: "Occurrence data overview for SDM"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 12
    fig-height: 8
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(ggplot2)
library(knitr)
library(scales)
library(rnaturalearth)
library(sf)
library(sampbias)
library(rnaturalearth)
library(dplyr)
library(sampbias)


# Suppress warnings for cleaner output
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Load data
db <- readRDS(here::here("data/db_modelling_strict.rds"))
```

# Dataset overview

```{r overview-table}
summary_stats <- tibble(
  Metric = c(
    "Total Records",
    "Records with Valid Coordinates",
    "Unique Species",
    "Unique Genera",
    "Unique Families",
    "Countries Represented",
    "Year Range",
    "Median Collection Year",
    "Data Source Groups"
  ),
  Value = c(
    format(nrow(db), big.mark = ","),
    format(sum(!is.na(db$decimalLatitude) & !is.na(db$decimalLongitude)), big.mark = ","),
    n_distinct(db$scientificName, na.rm = TRUE),
    n_distinct(db$genus, na.rm = TRUE),
    n_distinct(db$family, na.rm = TRUE),
    n_distinct(db$country, na.rm = TRUE),
    paste0(min(db$year, na.rm = TRUE), " – ", max(db$year, na.rm = TRUE)),
    round(median(as.numeric(db$year), na.rm = TRUE)),
    n_distinct(db$source_group, na.rm = TRUE)
  )
)

knitr::kable(summary_stats, caption = "Summary Statistics")
```

## Field completeness

```{r completeness}
key_fields <- c("decimalLatitude", "decimalLongitude", "year", "locality", 
                "minimumElevationInMeters", "family", "genus", "scientificName")

completeness <- tibble(
  Field = key_fields,
  Count = sapply(key_fields, function(col) sum(!is.na(db[[col]]))),
  Completeness_Percent = sapply(key_fields, function(col) {
    round(sum(!is.na(db[[col]])) / nrow(db) * 100, 1)
  })
) %>%
  arrange(desc(Completeness_Percent))

knitr::kable(completeness, caption = "Field Completeness")
```

## Rarity structure

>"On average, epiphytes have larger ranges than closely related terrestrial species, supporting hypotheses that epiphytism promotes dispersal. However, small ranges are prevalent in epiphyte-rich families regardless of lifeform. Notably, about half of epiphyte species are rare, indicating greater vulnerability than terrestrials. Epiphyte rarity is attributable to evolutionary history and shared traits rather than epiphytism itself." Svahnström et al., 2025. https://doi.org/10.1038/s41477-025-02022-9

>"Although epiphytism itself does not cause small ranges","the small ranges of epiphytes result from their concentration in a few species-rich lineages, where diversification and trait evolution have promoted
small ranges." 
And in conservation matters, "45% had EOO <20,000 km2, the threshold for
categorization as ‘vulnerable’ following Red List criterion B1"

SDM classic approach excluding rare species is not a good option for this group.

```{r brom-rarity-table1}
sp_occurrences <- db %>%
  filter(!is.na(scientificName)) %>%
  group_by(scientificName) %>%
  summarise(n_occ = n(), .groups = "drop")

rarity_bins <- sp_occurrences %>%
  mutate(
    bin = case_when(
      n_occ < 10 ~ "<10",
      n_occ >= 10 & n_occ < 50 ~ "10-50",
      n_occ >= 50 & n_occ < 100 ~ "50-100",
      n_occ >= 100 ~ ">100"
    )
  ) %>%
  count(bin) %>%
  mutate(bin = factor(bin, levels = c("<10", "10-50", "50-100", ">100"))) %>%
  arrange(bin)

# Extract values for inline text
rarity_rare <- rarity_bins %>% filter(bin == "<10") %>% pull(n)
rarity_common <- rarity_bins %>% filter(bin == ">100") %>% pull(n)
rarity_total <- sum(rarity_bins$n)

rarity_bins_table <- rarity_bins %>%
  rename("Occurrences per species" = bin, "Number of species" = n)

knitr::kable(rarity_bins_table, caption = "Distribution of species by occurrence count")
```

Among all species, rare species (fewer than 10 records) dominate numerically: `r rarity_rare` out of `r rarity_total` species. The tail of well-sampled species (\>100 records) contains only `r rarity_common` species, yet these represent a disproportionate fraction of the total records, illustrating the common-species-dominated sampling bias in occurrence data.

```{r rarity-histogram}
ggplot(sp_occurrences, aes(x = n_occ)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  scale_x_log10(
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  labs(
    title = "Distribution of occurrences per species (log scale)",
    x = "Records per species",
    y = "Number of species"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12, margin = margin(b = 8))
  )
```

### Species rarity: rare species (<10 and <20 records)

```{r rarity-species-histograms}
# Calculate occurrences per species
sp_occurrences <- db %>%
  filter(!is.na(scientificName)) %>%
  group_by(scientificName) %>%
  summarise(n_records = n(), .groups = "drop")

# Species with < 10 records
sp_less_10 <- sp_occurrences %>%
  filter(n_records < 10)

# Species with < 20 records
sp_less_20 <- sp_occurrences %>%
  filter(n_records < 20)

# Histogram for species with < 10 records
p1 <- ggplot(sp_less_10, aes(x = n_records)) +
  geom_histogram(bins = 9, fill = "steelblue", color = "white", alpha = 0.8) +
  labs(
    title = "Distribution of Records for Species with < 10 Occurrences",
    x = "Number of Records",
    y = "Number of Species"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 12))

# Histogram for species with < 20 records
p2 <- ggplot(sp_less_20, aes(x = n_records)) +
  geom_histogram(bins = 19, fill = "coral", color = "white", alpha = 0.8) +
  labs(
    title = "Distribution of Records for Species with < 20 Occurrences",
    x = "Number of Records",
    y = "Number of Species"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 12))

# Display both plots
p1
p2
```

There are `r nrow(sp_less_10)` species with fewer than 10 records and `r nrow(sp_less_20)` species with fewer than 20 records in the dataset. This substantial pool of rare species is a key consideration for SDM development, as they require specialized approaches beyond standard modeling techniques.

```{r rare-species-table}
# Get detailed info for species with < 20 records
rare_species_detail <- db %>%
  filter(!is.na(scientificName), !is.na(family), !is.na(genus)) %>%
  group_by(family, genus, scientificName) %>%
  summarise(
    n_records_total = n(),
    n_r_occurrences = sum(dataset == "Occurrence data"),
    n_r_communities = sum(dataset == "Community data"),
    .groups = "drop"
  ) %>%
  filter(n_records_total < 20) %>%
  arrange(desc(n_records_total))

knitr::kable(rare_species_detail[c(1:10),], 
             caption = "Species with < 20 records: Family, genus, species, and record counts by data type",
             col.names = c("Family", "Genus", "Species", "Total Records",
                           "Occurrences_source", "Communities_source"))

dim(rare_species_detail)
# Count species with at least one community record
rare_species_detail %>%
   filter(n_r_communities > 0) %>%
   nrow()
```

```{r rare-species-table10}
# Get detailed info for species with < 20 records
rare_species_detail1 <- db %>%
  filter(!is.na(scientificName), !is.na(family), !is.na(genus)) %>%
  group_by(family, genus, scientificName) %>%
  summarise(
    n_records_total = n(),
    n_r_occurrences = sum(dataset == "Occurrence data"),
    n_r_communities = sum(dataset == "Community data"),
    .groups = "drop"
  ) %>%
  filter(n_records_total < 10) %>%
  arrange(desc(n_records_total))

knitr::kable(rare_species_detail1[c(1:10),], 
             caption = "Species with < 10 records: Family, genus, species, and record counts by data type",
             col.names = c("Family", "Genus", "Species", "Total Records",
                           "Occurrences_source", "Communities_source"))

dim(rare_species_detail1)

# Count species with at least one community record
rare_species_detail1 %>%
   filter(n_r_communities > 0) %>%
   nrow()
```

## Geographic coverage

### Occurrence point distribution

```{r geographic-coverage}
db_coords <- db %>%
  filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) %>%
  filter(decimalLatitude >= -90 & decimalLatitude <= 90,
         decimalLongitude >= -180 & decimalLongitude <= 180)

# Sample for visualization
set.seed(4821)
db_sample <- db_coords %>% slice_sample(n = min(50000, nrow(db_coords)))

ggplot(db_sample, aes(x = decimalLongitude, y = decimalLatitude, color = source_group)) +
  geom_point(size = 1, alpha = 0.4) +
  coord_fixed() +
  scale_color_viridis_d(option = "turbo", name = "Source Group") +
  theme_minimal() +
  labs(title = "Geographic Distribution of Occurrence Points",
       subtitle = sprintf("Sample of %d from %d total points with valid coordinates", 
                          nrow(db_sample), nrow(db_coords)),
       x = "Longitude", y = "Latitude") +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 12))
```

```{r}
# Aggregate records into 0.5° grid cells for fast rendering
grid_counts <- db |>
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) |>
  mutate(
    lon_bin = round(decimalLongitude / 0.5) * 0.5,
    lat_bin = round(decimalLatitude  / 0.5) * 0.5
  ) |>
  count(lon_bin, lat_bin)

nt <- st_read("../layers/Neotropics_Morrone/Neotropic_dissolved.shp")
nt_simple <- st_simplify(nt, dTolerance = 0.05)

ggplot() +
  geom_sf(data = nt_simple, fill = "grey93", colour = "grey65", linewidth = 0.2) +
  geom_tile(
    data = grid_counts,
    aes(x = lon_bin, y = lat_bin, fill = n),
    width = 0.5, height = 0.5
  ) +
  scale_fill_viridis_c(
    option    = "magma",
    direction = -1,
    trans     = "log10",
    labels    = scales::comma,
    name      = "Records\n(log\u2081\u2080)"
  ) +
  coord_sf(xlim = c(-115, -30), ylim = c(-35, 25), expand = FALSE) +
  labs(
    title    = "EpIG 2.0 \u2014 occurrence density (0.5\u00b0 grid)",
    subtitle = paste0(
      format(sum(grid_counts$n), big.mark = ","),
      " georeferenced records \u00b7 ",
      format(nrow(grid_counts), big.mark = ","), " grid cells"
    ),
    x = NULL, y = NULL
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title        = element_text(face = "bold", margin = margin(b = 4)),
    plot.subtitle     = element_text(colour = "grey40"),
    legend.position   = "right",
    legend.key.height = unit(1.2, "cm"),
    plot.margin       = margin(8, 8, 8, 8)
  )
```

### Species Richness Grid

```{r richness-grid}
richness_grid <- db_coords %>%
  filter(!is.na(scientificName)) %>%
  mutate(
    lat_cell = floor(decimalLatitude),
    lon_cell = floor(decimalLongitude)
  ) %>%
  group_by(lat_cell, lon_cell) %>%
  summarise(
    n_species = n_distinct(scientificName),
    n_records = n(),
    .groups = "drop"
  ) %>%
  filter(!is.na(lat_cell) & !is.na(lon_cell))

ggplot(richness_grid, aes(x = lon_cell, y = lat_cell, fill = n_species)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_viridis_c(option = "viridis", direction = -1, name = "Unique Species") +
  theme_minimal() +
  labs(title = "Species Richness by 1° Grid Cell",
       x = "Longitude", y = "Latitude") +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "bottom")

richness_summary <- richness_grid %>%
  summarise(
    Mean_Species_per_Cell = round(mean(n_species), 1),
    Max_Species_in_Cell = max(n_species),
    Median_Species_per_Cell = median(n_species),
    Total_Grid_Cells = n()
  )

knitr::kable(richness_summary, caption = "Species Richness Summary Statistics")
```

## Sampling bias

```{r sampbias-setup}
# Prepare data for sampbias
# Important: sampbias needs a species column, not only coordinates
db_coords_sb <- db %>%
  filter(
    !is.na(decimalLatitude),
    !is.na(decimalLongitude),
    decimalLatitude >= -90,
    decimalLatitude <= 90,
    decimalLongitude >= -180,
    decimalLongitude <= 180,
    !is.na(scientificName),
    scientificName != ""
  ) %>%
  transmute(
    species = scientificName,
    decimalLongitude = as.numeric(decimalLongitude),
    decimalLatitude  = as.numeric(decimalLatitude)
  )

# Calculate sampling bias
sampbias_result <- calculate_bias(
  x = db_coords_sb,
  res = 2,
  terrestrial = TRUE
)

# See summary
summary(sampbias_result)

# Plot posterior estimates of bias weights
plot(sampbias_result)

# 4. Proyectar el sesgo al espacio
sampbias_map <- project_bias(sampbias_result)

# 5. Mapa del sesgo en escala log10
map_bias(
  sampbias_map,
  type = "diff_to_max"
)

map_bias(
  sampbias_map,
  type = "sampling_rate"
)
```

The sampling bias magnitude plot reveals geographic clustering of collection effort, with particular concentration in accessible regions and near research institutions. Regions with high bias (warmer colors) indicate non-random sampling intensity that must be accounted for in downstream SDM applications.

## Taxonomic Composition

## Family concentration

```{r family-concentration}
# Calculate species concentration across families
family_species <- db %>%
  filter(!is.na(family), !is.na(scientificName)) %>%
  group_by(family) %>%
  summarise(n_species = n_distinct(scientificName), .groups = "drop") %>%
  arrange(desc(n_species))

# Total families and total species
total_families <- nrow(family_species)
total_species_all <- sum(family_species$n_species)

# Top 6 families
top_10_families <- family_species %>%
  head(10) %>%
  mutate(pct_of_total = round(100 * n_species / total_species_all, 2))

# Concentration in top 6
concentration_top_10 <- round(100 * sum(top_10_families$n_species) / total_species_all, 1)

# Create summary text with inline values
species_top_10_table <- top_10_families %>%
  rename(Family = family, Species = n_species, `% of Total` = pct_of_total) %>%
  select(Family, Species, `% of Total`)

knitr::kable(species_top_10_table, caption = "Top 10 Families by Species Count")
```

In this dataset, `r concentration_top_10`% of all epiphyte species are concentrated in just the top 10 families. There are `r total_families` families total, but species diversity is heavily skewed: the top family has `r family_species %>% slice(1) %>% pull(n_species)` species, while the top 10 families account for `r sum(top_10_families$n_species)` species combined.

### Top Families

```{r families}
family_counts <- db %>%
  filter(!is.na(family)) %>%
  count(family) %>%
  arrange(desc(n)) %>%
  head(15) %>%
  mutate(pct = round(100 * n / sum(n), 1))

ggplot(family_counts, aes(x = reorder(family, n), y = n, fill = n)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(direction = -1, option = "viridis") +
  theme_minimal() +
  labs(title = "Top 15 Plant Families by Record Count",
       x = "Family", y = "Number of Records", fill = "Records") +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "bottom")
```

```{r family-species-table}
family_richness <- db %>%
  filter(!is.na(family), !is.na(genus), !is.na(scientificName)) %>%
  group_by(family) %>%
  summarise(
    Records = n(),
    Genus = n_distinct(genus),
    Species = n_distinct(scientificName),
    .groups = "drop"
  ) %>%
  arrange(desc(Records)) %>%
  head(15)

knitr::kable(family_richness, caption = "Top 15 Families: Records and Species Count")
```

### Top Genera

```{r genera}
genus_counts <- db %>%
  filter(!is.na(genus)) %>%
  count(genus) %>%
  arrange(desc(n)) %>%
  head(15)

ggplot(genus_counts, aes(x = reorder(genus, n), y = n, fill = n)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(direction = -1, option = "viridis") +
  theme_minimal() +
  labs(title = "Top 15 Genera by Record Count",
       x = "Genus", y = "Number of Records", fill = "Records") +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "bottom")
```

```{r genera-species-richness}
genus_richness <- db %>%
  filter(!is.na(genus), !is.na(scientificName)) %>%
  group_by(genus) %>%
  summarise(n_species = n_distinct(scientificName), .groups = "drop") %>%
  arrange(desc(n_species)) %>%
  head(15)

ggplot(genus_richness, aes(x = reorder(genus, n_species), y = n_species, fill = n_species)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_c(direction = -1, option = "plasma") +
  theme_minimal() +
  labs(title = "Top 15 Genera by Species Richness",
       x = "Genus", y = "Number of Species", fill = "Species") +
  theme(plot.title = element_text(face = "bold", size = 12),
        legend.position = "bottom")
```

```{r genera-table}
# Combine records and species counts for top genera
genus_combined <- db %>%
  filter(!is.na(genus), !is.na(scientificName)) %>%
  group_by(genus) %>%
  summarise(
    Records = n(),
    Species = n_distinct(scientificName),
    .groups = "drop"
  ) %>%
  arrange(desc(Species)) %>%
  head(15)

knitr::kable(genus_combined, caption = "Top 15 Genera: Records and Species Count")
```

### Geographic distribution of top genera

```{r top-genera-map}
# Get top 6 genera by species richness
top_6_genera <- db %>%
  filter(!is.na(genus), !is.na(scientificName)) %>%
  group_by(genus) %>%
  summarise(n_species = n_distinct(scientificName), .groups = "drop") %>%
  arrange(desc(n_species)) %>%
  head(6) %>%
  pull(genus)

# Filter data for top 6 genera with valid coordinates
map_top_genera <- db %>%
  filter(genus %in% top_6_genera) %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
  filter(decimalLatitude >= -90 & decimalLatitude <= 90,
         decimalLongitude >= -180 & decimalLongitude <= 180)

# Sample for cleaner visualization if needed
set.seed(8374)
map_top_genera_plot <- map_top_genera %>%
  slice_sample(n = min(25000, nrow(map_top_genera)))

# Get world map
world_map <- ne_countries(scale = "medium", returnclass = "sf")

# Create map with Tillandsia assigned the brightest color
genus_colors <- c(
  "Tillandsia" = "#FFD700",    # Bright gold/yellow (clearest)
  "Lepanthes" = "#FF6B6B",   # Coral red
  "Stelis" = "#A52A2A",      # Teal
  "Anthurium" = "#95E1D3",       # Light mint
  "Epidendrum" = "#6C5CE7",        # Purple
  "Peperomia" = "#808000"     # Olive
)

ggplot() +
  geom_sf(data = world_map, fill = "grey95", color = "grey70", linewidth = 0.3) +
  geom_point(
    data = map_top_genera_plot,
    aes(x = decimalLongitude, y = decimalLatitude, color = genus),
    size = 1.5, alpha = 0.6
  ) +
  facet_wrap(~genus) +
  scale_color_manual(values = genus_colors, name = "Genus") +
  coord_sf(xlim = c(-115, -30), ylim = c(-35, 25), expand = FALSE) +
  labs(
    title = "Geographic distribution of the top 6 genera (by species richness)",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12, margin = margin(b = 8)),
    legend.position = "right"
  )
```

```{r}
# Get lower 6 genera by species richness
top_6_genera <- db %>%
  filter(!is.na(genus), !is.na(scientificName)) %>%
  group_by(genus) %>%
  summarise(n_species = n_distinct(scientificName), .groups = "drop") %>%
  arrange(desc(n_species)) %>%
  head(6) %>%
  pull(genus)
```

## Temporal distribution

### Collection Effort Over Time

current species ranges? though historical records are also valuable for understanding range dynamics

```{r temporal}
timeline <- db %>%
  filter(!is.na(year)) %>%
  mutate(year = as.numeric(year)) %>%
  filter(year >= 1900 & year <= 2026) %>%
  count(year) %>%
  arrange(year)

ggplot(timeline, aes(x = year, y = n)) +
  geom_area(fill = "steelblue", alpha = 0.6) +
  geom_line(color = "steelblue", size = 0.8) +
  theme_minimal() +
  labs(title = "Collection Effort Over Time (1900–2026)",
       x = "Year", y = "Number of Records") +
  theme(plot.title = element_text(face = "bold", size = 12)) +
  scale_x_continuous(breaks = seq(1900, 2026, 25))

# Summary statistics
year_summary <- db %>%
  filter(!is.na(year)) %>%
  mutate(year = as.numeric(year)) %>%
  summarise(
    Records_Pre_1980 = sum(year < 1980),
    Records_1980_2000 = sum(year >= 1980 & year < 2000),
    Records_2000_2020 = sum(year >= 2000 & year < 2020),
    Records_Post_2020 = sum(year >= 2020)
  ) %>%
  pivot_longer(everything(), names_to = "Period", values_to = "Records")

knitr::kable(year_summary, caption = "Records by Decade")
```

# Pilot

```{r}
# Species rarity distribution
species_rarity <- db %>%
  group_by(scientificName) %>%
  summarise(n_records = n(), .groups = 'drop') %>%
  mutate(
    record_class = case_when(
      n_records == 1 ~ "1 record",
      n_records < 5 ~ "2-4 records",
      n_records < 10 ~ "5-9 records",
      n_records < 20 ~ "10-19 records",
      n_records < 30 ~ "20-29 records",
      n_records < 50 ~ "30-49 records",
      n_records < 100 ~ "50-99 records",
      TRUE ~ "100+ records"
    )
  ) %>%
  count(record_class) %>%
  rename(`Record Class` = record_class, `Number of Species` = n) %>%
  mutate(`Record Class` = factor(`Record Class`, 
    levels = c("1 record", "2-4 records", "5-9 records", "10-19 records",
               "20-29 records", "30-49 records", "50-99 records", "100+ records"))) %>%
  arrange(`Record Class`)

knitr::kable(species_rarity, caption = "Distribution of Species by Number of Records")

# Genus rarity distribution
genus_rarity <- db %>%
  group_by(genus) %>%
  summarise(n_records = n(), .groups = 'drop') %>%
  mutate(
    record_class = case_when(
      n_records == 1 ~ "1 record",
      n_records < 5 ~ "2-4 records",
      n_records < 10 ~ "5-9 records",
      n_records < 20 ~ "10-19 records",
      n_records < 30 ~ "20-29 records",
      n_records < 50 ~ "30-49 records",
      n_records < 100 ~ "50-99 records",
      TRUE ~ "100+ records"
    )
  ) %>%
  count(record_class) %>%
  rename(`Record Class` = record_class, `Number of Genera` = n) %>%
  mutate(`Record Class` = factor(`Record Class`, 
    levels = c("1 record", "2-4 records", "5-9 records", "10-19 records",
               "20-29 records", "30-49 records", "50-99 records", "100+ records"))) %>%
  arrange(`Record Class`)

knitr::kable(genus_rarity, caption = "Distribution of Genera by Number of Records")

# Filter species with >= 5 records
species_5plus <- db %>%
  group_by(scientificName) %>%
  filter(n() >= 5) %>%
  ungroup()

# Count remaining species and families
species_count_filtered <- species_5plus %>%
  pull(scientificName) %>%
  n_distinct()

family_count_filtered <- species_5plus %>%
  pull(family) %>%
  n_distinct()

records_filtered <- nrow(species_5plus)

summary_after_filter <- tibble(
  Metric = c("Number of Records", "Number of Species", "Number of Families"),
  Count = c(records_filtered, species_count_filtered, family_count_filtered)
)

knitr::kable(summary_after_filter, caption = "Summary After Removing Species with < 5 Records")
```


## GENUS LEVEL APPROACH: pilot genus: Epidendrum and Andinia

Andinia has 27 species
Bulbophyllum 71
Epidendrum 1063

```{r}
# Genus restricted to the Andes, widespread, widespread+diverse
target <- c("Andinia", "Bulbophyllum", "Epidendrum")

# Filter data for top 6 genera with valid coordinates
map_target_genera <- db %>%
  filter(genus %in% target) %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
  filter(decimalLatitude >= -90 & decimalLatitude <= 90,
         decimalLongitude >= -180 & decimalLongitude <= 180)

# Sample for cleaner visualization if needed
set.seed(8374)
map_target_genera <- map_target_genera %>%
  slice_sample(n = min(25000, nrow(map_target_genera)))

# Get world map
world_map <- ne_countries(scale = "medium", returnclass = "sf")

# Create map with Tillandsia assigned the brightest color
genus_colors <- c(
  "Epidendrum" = "#FFD700",    # Bright gold/yellow (clearest)
  "Bulbophyllum" = "#FF6B6B",   # Coral red
  "Andinia" = "#6C5CE7"        # Purple
)

ggplot() +
  geom_sf(data = world_map, fill = "grey95", color = "grey70", linewidth = 0.3) +
  geom_point(
    data = map_target_genera,
    aes(x = decimalLongitude, y = decimalLatitude, color = genus),
    size = 1.5, alpha = 0.6
  ) +
  facet_wrap(~genus) +
  scale_color_manual(values = genus_colors, name = "Genus") +
  coord_sf(xlim = c(-115, -30), ylim = c(-35, 25), expand = FALSE) +
  labs(
    title = "Geographic distribution of target genera",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12, margin = margin(b = 8)),
    legend.position = "right"
  )


db_genus_sf <- db %>%
  filter(genus == target)

genus_summary <- db_genus_sf %>%
  group_by(genus) %>%
  summarise(
    Records = n(),
    Species = n_distinct(scientificName),
    .groups = "drop"
  ) %>%
  arrange(desc(Records))

# Rarity structure
species_occurrences_ <- db_genus_sf %>%
  group_by(genus, scientificName) %>%
  summarise(n_occ = n(), .groups = "drop")

rarity_bins <- species_occurrences_ %>%
  mutate(
    bin = case_when(
      n_occ < 10 ~ "<10",
      n_occ >= 10 & n_occ < 50 ~ "10-50",
      n_occ >= 50 & n_occ < 100 ~ "50-100",
      n_occ >= 100 ~ ">100"
    )
  ) %>%
  count(bin) %>%
  mutate(bin = factor(bin, levels = c("<10", "10-50", "50-100", ">100"))) %>%
  arrange(bin)

# Extract values for inline text
rarity_rare <- rarity_bins %>% filter(bin == "<10") %>% pull(n)
rarity_common <- rarity_bins %>% filter(bin == ">100") %>% pull(n)
rarity_total <- sum(rarity_bins$n)

rarity_bins_table <- rarity_bins %>%
  rename("Occurrences per species" = bin, "Number of species" = n)

knitr::kable(rarity_bins_table, caption = "Distribution of species by occurrence count")
```

```{r}

```

## SPECIES LEVEL APPROACH: pilot family: Bromeliaceae

```{r brom-setup, include=FALSE}
library(tidyverse)

# Subfamily assignments
subfamily_map <- tribble(
  ~genus, ~subfamily,
  "Tillandsia", "Tillandsioideae",
  "Connellia", "Tillandsioideae",
  "Glomeraria", "Tillandsioideae",
  "Racinaea", "Tillandsioideae",
  "Alcantarea", "Tillandsioideae",
  "Vriesea", "Bromelioideae",
  "Nidularium", "Bromelioideae",
  "Aechmea", "Bromelioideae",
  "Billbergia", "Bromelioideae",
  "Hohenbergia", "Bromelioideae",
  "Wittrockia", "Bromelioideae",
  "Dyckia", "Pitcairnioideae",
  "Puya", "Pitcairnioideae",
  "Deuterocohnia", "Pitcairnioideae",
  "Abromeitiella", "Pitcairnioideae"
)

db_brom_sf <- db %>%
  filter(family == "Bromeliaceae") %>%
  left_join(subfamily_map, by = "genus") %>%
  mutate(subfamily = ifelse(is.na(subfamily), "Unknown", subfamily))

# Summary statistics
brom_records <- nrow(db_brom_sf)
brom_species <- n_distinct(db_brom_sf$scientificName)
total_records <- nrow(db)
total_species <- n_distinct(db$scientificName)

# Subfamily summaries
subfamily_summary <- db_brom_sf %>%
  filter(!is.na(subfamily) & subfamily != "Unknown") %>%
  group_by(subfamily) %>%
  summarise(
    Records = n(),
    Species = n_distinct(scientificName),
    Genera = n_distinct(genus),
    .groups = "drop"
  ) %>%
  arrange(desc(Records))

# Rarity structure
species_occurrences <- db_brom_sf %>%
  filter(!is.na(scientificName)) %>%
  group_by(scientificName) %>%
  summarise(n_occ = n(), .groups = "drop")

rarity_bins <- species_occurrences %>%
  mutate(
    bin = case_when(
      n_occ < 10 ~ "<10",
      n_occ >= 10 & n_occ < 50 ~ "10-50",
      n_occ >= 50 & n_occ < 100 ~ "50-100",
      n_occ >= 100 ~ ">100"
    )
  ) %>%
  count(bin) %>%
  mutate(bin = factor(bin, levels = c("<10", "10-50", "50-100", ">100"))) %>%
  arrange(bin)

# Extract values for inline text
rarity_rare <- rarity_bins %>% filter(bin == "<10") %>% pull(n)
rarity_common <- rarity_bins %>% filter(bin == ">100") %>% pull(n)
rarity_total <- sum(rarity_bins$n)
```

Bromeliaceae is the second-largest plant family in this vascular epiphyte dataset, representing `r format(brom_records, big.mark = ",")` records (`r round(100 * brom_records / total_records, 1)`%) and `r brom_species` species (`r round(100 * brom_species / total_species, 1)`% of all species). The family is near-exclusively distributed across the **Neotropics**, spanning the full geographic extent of the study region from Mexico to southern Brazil.

Within the epiphyte-only context of this dataset, Bromeliaceae is structured around two main subfamilies with distinctly different ecological strategies. Tillandsioideae, the atmospheric epiphytes, account for `r subfamily_summary %>% filter(subfamily == "Tillandsioideae") %>% pull(Records) %>% format(big.mark = ",")` records across `r subfamily_summary %>% filter(subfamily == "Tillandsioideae") %>% pull(Species)` species and occupy the broadest altitudinal and climatic range. Bromelioideae, the tank epiphytes, comprise `r subfamily_summary %>% filter(subfamily == "Bromelioideae") %>% pull(Records) %>% format(big.mark = ",")` records and `r subfamily_summary %>% filter(subfamily == "Bromelioideae") %>% pull(Species)` species, with stronger preference for humid forest habitats.

The family also exhibits the same rarity structure as the full dataset: a heavy-tailed distribution with a few dominant species and many rare ones. Among Bromeliaceae species, `r rarity_rare` have fewer than 10 occurrences, while only `r rarity_common` have over 100 records. The variation creates a natural test case for a bifurcated modeling approach: developing specialized methods for common species (where sufficient data exists for standard maximum entropy modeling) and rare species (where simpler or data-borrowing approaches may be needed).

The strong phylogenetic structure within Bromeliaceae—two well-defined subfamilies with clear ecological associations—also allows us to test whether modeling coherence and prediction quality improve when conditioning on higher taxonomic levels.

```{r brom-dataset-comparison}
# Dataset comparison
comparison <- tibble(
  Group = c("Full Dataset", "Bromeliaceae"),
  Records = c(total_records, brom_records),
  Species = c(total_species, brom_species),
  `Records (%)` = c(100, round(100 * brom_records / total_records, 1)),
  `Species (%)` = c(100, round(100 * brom_species / total_species, 1))
)

knitr::kable(comparison, caption = "Bromeliaceae contribution to the full dataset")
```

### Subfamily composition

```{r brom-subfamily}
subfamily_summary <- db_brom_sf %>%
  filter(!is.na(subfamily) & subfamily != "Unknown") %>%
  group_by(subfamily) %>%
  summarise(
    Records = n(),
    Species = n_distinct(scientificName),
    Genera = n_distinct(genus),
    .groups = "drop"
  ) %>%
  arrange(desc(Records))

knitr::kable(subfamily_summary, caption = "Records, species, and genera per Bromeliaceae subfamily")
```

### Rarity structure

```{r brom-rarity-table}
rarity_bins_table <- rarity_bins %>%
  rename("Occurrences per species" = bin, "Number of species" = n)

knitr::kable(rarity_bins_table, caption = "Distribution of species by occurrence count")
```

Among Bromeliaceae species, rare species (fewer than 10 records) dominate numerically: `r rarity_rare` out of `r rarity_total` species. The tail of well-sampled species (\>100 records) contains only `r rarity_common` species, yet these represent a disproportionate fraction of the total records, illustrating the common-species-dominated sampling bias in occurrence data.

```{r brom-rarity-histogram}
ggplot(species_occurrences, aes(x = n_occ)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white", alpha = 0.8) +
  scale_x_log10(
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  labs(
    title = "Distribution of occurrences per Bromeliaceae species (log scale)",
    x = "Records per species",
    y = "Number of species"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12, margin = margin(b = 8))
  )
```

### Top genera

```{r brom-genera}
top_genera <- db_brom_sf %>%
  filter(!is.na(genus)) %>%
  count(genus) %>%
  arrange(desc(n)) %>%
  head(10) %>%
  rename(Genus = genus, Records = n)

knitr::kable(top_genera, caption = "Top 10 genera in Bromeliaceae")
```

The distribution is predictably skewed toward *Tillandsia* (`r (top_genera %>% filter(Genus == "Tillandsia") %>% pull(Records)) %>% format(big.mark = ",")` records), but the next tier of genera—*Aechmea*, *Guzmania*, and *Vriesea*—provides sufficient sample sizes for independent model development and cross-validation.

### Geographic distribution

```{r brom-map}
library(rnaturalearth)
library(sf)

map_data <- db_brom_sf %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude), 
         subfamily != "Unknown") %>%
  filter(decimalLatitude >= -90 & decimalLatitude <= 90,
         decimalLongitude >= -180 & decimalLongitude <= 180)

set.seed(7423)
map_data_plot <- map_data %>%
  slice_sample(n = min(30000, nrow(map_data)))

world <- ne_countries(scale = "medium", returnclass = "sf")

ggplot() +
  geom_sf(data = world, fill = "grey95", color = "grey70", linewidth = 0.3) +
  geom_point(
    data = map_data_plot,
    aes(x = decimalLongitude, y = decimalLatitude, color = subfamily),
    size = 1.5, alpha = 0.5
  ) +
  scale_color_manual(
    values = c("Tillandsioideae" = "#440154", "Bromelioideae" = "#FDE725"),
    name = "Subfamily"
  ) +
  coord_sf(xlim = c(-115, -30), ylim = c(-35, 25), expand = FALSE) +
  labs(
    title = "Bromeliaceae occurrence distribution by subfamily",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12, margin = margin(b = 8)),
    legend.position = "right"
  )
```

The two subfamilies show clear geographic complementarity. Tillandsioideae (purple) dominates the northern and western regions, extending into Mexico and along the Andes, while Bromelioideae (yellow) is concentrated in the humid Atlantic forest regions of southeastern Brazil. This geographic structure, combined with their ecological differences, makes the subfamily distinction meaningful for modeling.

# Hotspots within the Neotropics

```{r}
library(sf)
library(dplyr)

# Load hotspots shapefile
hotspots <- st_read("../layers/hotspots_2016_1/hotspots_2016_1.shp", quiet = TRUE)
hotspots <- hotspots %>% filter(Type == "hotspot area")

# Load Neotropics boundary
neotropics <- st_read("../layers/Neotropics_Morrone/Neotropic_dissolved.shp", quiet = TRUE)

# Extract attributes without geometry (shapefile has topology issues)
hotspots_attrs <- st_drop_geometry(hotspots) 

# Identify hotspots in Neotropics by name pattern matching
neotropical_hotspots_keywords <- c(
  "Atlantic Forest",
  "Caribbean Islands",
  "Cerrado",
  "Madrean Pine-Oak Woodlands",
  # "Chilean Winter Rainfall and Valdivian Forests",
  "Mesoamerica",
  "Tropical Andes",
  "Tumbes-Choco-Magdalena"
)

hotspots_neotropics <- hotspots_attrs %>%
  filter(grepl(paste(neotropical_hotspots_keywords, collapse = "|"), NAME, ignore.case = TRUE))
# hotspots_neotropics %>% unique(NAME)

# Report results
cat("Total hotspots loaded:", nrow(hotspots), "\n")
cat("Hotspots within Neotropics:", nrow(hotspots_neotropics), "\n")

# Display hotspots within Neotropics
if (nrow(hotspots_neotropics) > 0) {
  cat("\nBiodiversity Hotspots in the Neotropics:\n")
  print(hotspots_neotropics)
  
  cat("\nSummary:\n")
  hotspot_types <- hotspots_neotropics %>%
    count(Type)
  print(hotspot_types)
  
  cat("\nUnique hotspots (excluding outer limits):\n")
  unique_hotspots <- hotspots_neotropics %>%
    filter(Type == "hotspot area") %>%
    pull(NAME)
  print(unique_hotspots)
} else {
  cat("No hotspots found in Neotropics\n")
}
```

# CHANGE versus previous version: 5 families

```{r}
library(tidyverse)
library(sf)
library(ggplot2)
library(hexbin)

glimpse(db)

# Load the previous database version
load("../data/database_final_2021.rda")
db_prev <- occs_final
db_prev <- db_prev %>% 
  mutate(scientificName = species)
db_prev_matched <- readRDS("../data/db_prev_matched.rds")

table(db_prev_matched$match_code)


# Create filtered db with target families only
db_filtered <- db %>%
  filter(family %in% c("Orchidaceae", "Bromeliaceae", "Araceae", "Piperaceae", "Polypodiaceae"))


# Detailed family breakdown for previous data
family_breakdown1 <- db_prev %>%
  group_by(family) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(scientificName),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_records))

cat("\nFamily breakdown for previous dataset:\n")
print(family_breakdown1)

# Detailed family breakdown for current data
family_breakdown2 <- db_filtered %>%
  group_by(family) %>%
  summarise(
    n_records = n(),
    n_species = n_distinct(scientificName),
    .groups = 'drop'
  ) %>%
  arrange(desc(n_records))

cat("\nFamily breakdown for current dataset:\n")
print(family_breakdown2)
```

♣ Prev dataset: EpIG-Backbone comparing

```{r}
library(dplyr)
library(ggplot2)

# --- 1. Define match groups ---
db_chk <- db_prev_matched %>%
  mutate(
    match_group = case_when(
      match_code == 1 ~ "match_1_exact",
      match_code == 2 ~ "match_2_fuzzy",
      match_code == 3 ~ "match_3_loose",
      is.na(match_code) | match_code == 0 ~ "no_match",
      TRUE ~ "other"
    ),
    lost_flag = ifelse(match_code == 1, "kept", "lost")
  )

# --- 2. GLOBAL SUMMARY (records) ---
summary_records <- db_chk %>%
  count(match_group) %>%
  mutate(prop = n / sum(n))

# --- 3. SPECIES LOSS ---
species_loss <- db_chk %>%
  group_by(scientificName) %>%
  summarise(
    total_records = n(),
    n_match1 = sum(match_code == 1, na.rm = TRUE),
    n_lost = sum(match_code != 1 | is.na(match_code)),
    prop_lost = n_lost / total_records,
    .groups = "drop"
  )

species_completely_lost <- species_loss %>%
  filter(n_match1 == 0)

# --- 4. FAMILY LEVEL (records + species) ---
loss_by_family <- db_chk %>%
  group_by(family) %>%
  summarise(
    total_records = n(),
    lost_records = sum(match_code != 1 | is.na(match_code)),
    prop_records_lost = lost_records / total_records,
    n_species_total = n_distinct(scientificName),
    .groups = "drop"
  ) %>%
  left_join(
    db_chk %>%
      group_by(family, scientificName) %>%
      summarise(
        has_match1 = any(match_code == 1, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      group_by(family) %>%
      summarise(
        n_species_lost = sum(!has_match1),
        .groups = "drop"
      ),
    by = "family"
  ) %>%
  mutate(
    prop_species_lost = n_species_lost / n_species_total
  ) %>%
  arrange(desc(lost_records))

# --- 5. GENUS LEVEL (records + species) ---
loss_by_genus <- db_chk %>%
  group_by(genus) %>%
  summarise(
    total_records = n(),
    lost_records = sum(match_code != 1 | is.na(match_code)),
    prop_records_lost = lost_records / total_records,
    n_species_total = n_distinct(scientificName),
    .groups = "drop"
  ) %>%
  left_join(
    db_chk %>%
      group_by(genus, scientificName) %>%
      summarise(
        has_match1 = any(match_code == 1, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      group_by(genus) %>%
      summarise(
        n_species_lost = sum(!has_match1),
        .groups = "drop"
      ),
    by = "genus"
  ) %>%
  mutate(
    prop_species_lost = n_species_lost / n_species_total
  ) %>%
  arrange(desc(lost_records))

# --- 6. SPATIAL PATTERN ---
map_data <- db_chk %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
  mutate(match_binary = ifelse(match_code == 1, "match_1", "non_match_1"))

ggplot(map_data, aes(x = decimalLongitude, y = decimalLatitude)) +
  geom_point(alpha = 0.4, size = 0.7) +
  facet_wrap(~match_binary) +
  theme_minimal()


db_chk <- db_prev_matched %>%
  mutate(
    lost = is.na(match_code) | match_code != 1
  )

# Species-level status
species_status <- db_chk %>%
  group_by(family, genus, scientificName) %>%
  summarise(
    total_records = n(),
    lost_records = sum(lost),
    kept_records = sum(!lost),
    completely_lost = all(lost),
    .groups = "drop"
  )

# Family table with lost species names
loss_by_family <- species_status %>%
  group_by(family) %>%
  summarise(
    total_records = sum(total_records),
    lost_records = sum(lost_records),
    prop_records_lost = lost_records / total_records,
    n_species_total = n_distinct(scientificName),
    n_species_lost = sum(completely_lost),
    prop_species_lost = n_species_lost / n_species_total,
    species_lost = paste(scientificName[completely_lost], collapse = "; "),
    .groups = "drop"
  ) %>%
  arrange(desc(lost_records))

# Genus table with lost species names
loss_by_genus <- species_status %>%
  group_by(family, genus) %>%
  summarise(
    total_records = sum(total_records),
    lost_records = sum(lost_records),
    prop_records_lost = lost_records / total_records,
    n_species_total = n_distinct(scientificName),
    n_species_lost = sum(completely_lost),
    prop_species_lost = n_species_lost / n_species_total,
    species_lost = paste(scientificName[completely_lost], collapse = "; "),
    .groups = "drop"
  ) %>%
  arrange(desc(lost_records))

head(loss_by_family, 20)
head(loss_by_genus, 20)

# --- OUTPUTS ---
summary_records
head(species_completely_lost, 20)
head(loss_by_family, 20)
head(loss_by_genus, 20)
```

comparing with new dataset

```{r}
library(dplyr)
library(ggplot2)

# --- 1. Prepare datasets ---
old <- db_prev_matched %>%
  filter(match_code == 1) %>%
  select(scientificName, family, genus, decimalLatitude, decimalLongitude)

new <- db_filtered %>%
  filter(matchCodeEpig == "1") %>%  # should already be all 1
  select(scientificName, family, genus, decimalLatitude, decimalLongitude)

# --- 2. BASIC SIZE ---
summary_sizes <- tibble(
  dataset = c("old", "new"),
  n_records = c(nrow(old), nrow(new)),
  n_species = c(n_distinct(old$scientificName),
                n_distinct(new$scientificName))
)

# --- 3. SPECIES COMPARISON (THIS IS THE CORE) ---
spp_old <- old %>% distinct(scientificName)
spp_new <- new %>% distinct(scientificName)

species_lost <- spp_old %>%
  anti_join(spp_new, by = "scientificName")

species_gained <- spp_new %>%
  anti_join(spp_old, by = "scientificName")

species_shared <- inner_join(spp_old, spp_new, by = "scientificName")

species_summary <- tibble(
  metric = c("old", "new", "shared", "lost", "gained"),
  n = c(nrow(spp_old),
        nrow(spp_new),
        nrow(species_shared),
        nrow(species_lost),
        nrow(species_gained))
)

# --- 4. TAXONOMIC SIGNAL ---
loss_by_family <- old %>%
  semi_join(species_lost, by = "scientificName") %>%
  group_by(family) %>%
  summarise(
    lost_records = n(),
    lost_species = n_distinct(scientificName),
    .groups = "drop"
  ) %>%
  arrange(desc(lost_records))

gain_by_family <- new %>%
  semi_join(species_gained, by = "scientificName") %>%
  group_by(family) %>%
  summarise(
    gained_records = n(),
    gained_species = n_distinct(scientificName),
    .groups = "drop"
  ) %>%
  arrange(desc(gained_records))

# --- 5. SPATIAL COMPARISON ---
map_data <- bind_rows(
  old %>% mutate(dataset = "old"),
  new %>% mutate(dataset = "new")
) %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude))

ggplot(map_data, aes(decimalLongitude, decimalLatitude)) +
  geom_point(alpha = 0.3, size = 0.6) +
  facet_wrap(~dataset) +
  theme_minimal()

# --- OUTPUTS ---
summary_sizes
species_summary

head(species_lost, 20)
head(species_gained, 20)

head(loss_by_family, 20)
head(gain_by_family, 20)
```

# IUCN check to 08.05.2026

```{r}
install.packages("remotes")
remotes::install_github("IUCN-UK/iucnredlist")
library(iucnredlist)
library(stringr)
library(tibble)
library(purrr)
library(dplyr)

# token IUCN
api <- init_api("sDHQ3R1rXX5EkmVDNHoWRo43eE3jK8cmC9FU")

# dpecies list
species <- db_model %>%
  pull(scientificName) %>%
  str_trim() %>%
  str_squish() %>%
  word(1, 2) %>%
  unique()

check_iucn <- function(sp, api){

  out <- tryCatch({

    parts <- str_split(
      sp,
      "\\s+",
      simplify = TRUE
    )

    genus <- parts[1]
    species_epithet <- parts[2]

    x <- assessments_by_name(
      api = api,
      genus = genus,
      species = species_epithet
    )

    assessed <- nrow(x) > 0

    category <- if (assessed) {
      x$red_list_category_code[1]
    } else {
      NA_character_
    }

    latest_year <- if (assessed) {
      x$year_published[1]
    } else {
      NA_character_
    }

    tibble(
      species = sp,
      assessed_by_iucn = assessed,
      category = category,
      latest_assessment = latest_year,
      n_assessments = nrow(x)
    )

  }, error = function(e){

    tibble(
      species = sp,
      assessed_by_iucn = FALSE,
      category = NA_character_,
      latest_assessment = NA_character_,
      n_assessments = NA_integer_
    )

  })

  Sys.sleep(0.2)

  out
}

results <- map_dfr(
  species,
  check_iucn,
  api = api
)

view(results)
write.csv(results, "data/iucn_2026.csv")
```


# Plots to export 

```{r}
# library(tidyverse)
# library(sf)
# library(ggplot2)
# library(plotly)
# library(viridis)
# library(gridExtra)
# 
# # Create figures directory
# dir.create("figures", showWarnings = FALSE)
# 
# # ============================================================================
# # SETUP: Prepare data summaries for fast visualization
# # ============================================================================
# 
# # Basic checks
# cat("Total records:", nrow(db), "\n")
# cat("Records with valid coordinates:", sum(!is.na(db$decimalLatitude) & !is.na(db$decimalLongitude)), "\n")
# cat("Unique families:", n_distinct(db$family), "\n")
# cat("Unique genera:", n_distinct(db$genus), "\n")
# cat("Year range:", min(db$year, na.rm = TRUE), "-", max(db$year, na.rm = TRUE), "\n")
# 
# # ============================================================================
# # MAPS
# # ============================================================================
# 
# cat("\n=== CREATING MAPS ===\n")
# 
# # Prepare coordinates with valid data
# db_coords <- db %>%
#   filter(!is.na(decimalLatitude) & !is.na(decimalLongitude)) %>%
#   filter(decimalLatitude >= -90 & decimalLatitude <= 90,
#          decimalLongitude >= -180 & decimalLongitude <= 180)
# 
# cat("Valid coordinates:", nrow(db_coords), "\n")
# 
# # --- 1. DOT DENSITY MAP (sample 50K for performance)
# cat("Creating dot density map...\n")
# set.seed(4821)
# db_sample <- db_coords %>% slice_sample(n = min(50000, nrow(db_coords)))
# 
# p_density <- ggplot(db_sample, aes(x = decimalLongitude, y = decimalLatitude, color = source_group)) +
#   geom_point(size = 0.5, alpha = 0.3) +
#   coord_fixed() +
#   scale_color_viridis_d(option = "turbo") +
#   theme_minimal() +
#   labs(title = "Geographic Distribution of Occurrence Points",
#        subtitle = sprintf("Sample of %d from %d total points with valid coordinates", 
#                           nrow(db_sample), nrow(db_coords)),
#        x = "Longitude", y = "Latitude", color = "Source Group") +
#   theme(legend.position = "bottom",
#         plot.title = element_text(face = "bold", size = 14),
#         plot.subtitle = element_text(size = 11))
# 
# ggsave("figures/01_dot_density_map.png", p_density, width = 15, height = 10, dpi = 300)
# cat("✓ Saved: 01_dot_density_map.png\n")
# 
# # --- 2. COUNTRY CHOROPLETH (record counts by country)
# cat("Creating country choropleth...\n")
# 
# country_counts <- db %>%
#   filter(!is.na(country)) %>%
#   count(country) %>%
#   arrange(desc(n)) %>%
#   head(30)
# 
# p_country <- ggplot(country_counts, aes(x = reorder(country, n), y = n, fill = n)) +
#   geom_col() +
#   coord_flip() +
#   scale_fill_viridis_c(direction = -1) +
#   theme_minimal() +
#   labs(title = "Record Count by Country (Top 30)",
#        subtitle = "Brazil and Mexico dominate; clear Neotropical bias",
#        x = "Country", y = "Number of Records", fill = "Records") +
#   theme(plot.title = element_text(face = "bold", size = 14))
# 
# ggsave("figures/02_country_distribution.png", p_country, width = 12, height = 10, dpi = 300)
# cat("✓ Saved: 02_country_distribution.png\n")
# 
# # --- 3. SPECIES RICHNESS GRID (1° cells)
# cat("Creating species richness grid...\n")
# 
# richness_grid <- db_coords %>%
#   filter(!is.na(scientificName)) %>%
#   mutate(
#     lat_cell = floor(decimalLatitude),
#     lon_cell = floor(decimalLongitude)
#   ) %>%
#   group_by(lat_cell, lon_cell) %>%
#   summarise(
#     n_species = n_distinct(scientificName),
#     n_records = n(),
#     .groups = "drop"
#   ) %>%
#   filter(!is.na(lat_cell) & !is.na(lon_cell))
# 
# p_richness <- ggplot(richness_grid, aes(x = lon_cell, y = lat_cell, fill = n_species)) +
#   geom_tile() +
#   coord_fixed() +
#   scale_fill_viridis_c(option = "viridis", direction = -1) +
#   theme_minimal() +
#   labs(title = "Species Richness by 1° Grid Cell",
#        subtitle = "Ecologically meaningful diversity pattern",
#        x = "Longitude", y = "Latitude", fill = "Unique Species") +
#   theme(plot.title = element_text(face = "bold", size = 14))
# 
# ggsave("figures/03_species_richness_grid.png", p_richness, width = 15, height = 10, dpi = 300)
# cat("✓ Saved: 03_species_richness_grid.png\n")
# 
# # ============================================================================
# # TAXONOMIC VISUALIZATIONS
# # ============================================================================
# 
# cat("\n=== CREATING TAXONOMIC VISUALIZATIONS ===\n")
# 
# # --- 4. TREEMAP (Families by record count)
# cat("Creating treemap...\n")
# 
# family_counts <- db %>%
#   filter(!is.na(family)) %>%
#   count(family) %>%
#   arrange(desc(n)) %>%
#   mutate(pct = round(100 * n / sum(n), 1))
# 
# # Create treemap manually since treemapify may not be available
# p_treemap <- ggplot(family_counts, aes(area = n, fill = n, label = paste0(family, "\n", n))) +
#   geom_treemap() +
#   geom_treemap_text(fontface = "italic", colour = "white", place = "centre") +
#   scale_fill_viridis_c(direction = -1) +
#   theme(plot.title = element_text(face = "bold", size = 14)) +
#   labs(title = "Taxonomic Distribution: Plant Families",
#        subtitle = "Orchidaceae and Bromeliaceae dominate",
#        fill = "Record Count")
# 
# ggsave("figures/04_family_treemap.png", p_treemap, width = 14, height = 10, dpi = 300)
# cat("✓ Saved: 04_family_treemap.png\n")
# 
# # --- 5. TOP 20 GENERA (horizontal bar chart with order coloring)
# cat("Creating top 20 genera chart...\n")
# 
# genus_counts <- db %>%
#   filter(!is.na(genus) & !is.na(order)) %>%
#   count(genus, order) %>%
#   arrange(desc(n)) %>%
#   head(20)
# 
# p_genera <- ggplot(genus_counts, aes(x = reorder(genus, n), y = n, fill = order)) +
#   geom_col() +
#   coord_flip() +
#   scale_fill_viridis_d(option = "turbo") +
#   theme_minimal() +
#   labs(title = "Top 20 Most Recorded Genera",
#        x = "Genus", y = "Number of Records", fill = "Order") +
#   theme(plot.title = element_text(face = "bold", size = 14))
# 
# ggsave("figures/05_top_genera.png", p_genera, width = 12, height = 10, dpi = 300)
# cat("✓ Saved: 05_top_genera.png\n")
# 
# # --- 6. TOP 20 FAMILIES (horizontal bar chart)
# cat("Creating top 20 families chart...\n")
# 
# top_families <- family_counts %>% head(20)
# 
# p_families <- ggplot(top_families, aes(x = reorder(family, n), y = n, fill = n)) +
#   geom_col() +
#   coord_flip() +
#   scale_fill_viridis_c(direction = -1) +
#   theme_minimal() +
#   labs(title = "Top 20 Plant Families",
#        x = "Family", y = "Number of Records", fill = "Records") +
#   theme(plot.title = element_text(face = "bold", size = 14))
# 
# ggsave("figures/06_top_families.png", p_families, width = 12, height = 9, dpi = 300)
# cat("✓ Saved: 06_top_families.png\n")
# 
# # ============================================================================
# # TEMPORAL VISUALIZATIONS
# # ============================================================================
# 
# cat("\n=== CREATING TEMPORAL VISUALIZATIONS ===\n")
# 
# # --- 7. COLLECTION EFFORT TIMELINE (records per year)
# cat("Creating timeline...\n")
# 
# timeline <- db %>%
#   filter(!is.na(year)) %>%
#   mutate(year = as.numeric(year)) %>%
#   filter(year >= 1692 & year <= 2026) %>%
#   count(year) %>%
#   arrange(year)
# 
# p_timeline <- ggplot(timeline, aes(x = year, y = n)) +
#   geom_area(fill = "steelblue", alpha = 0.6) +
#   geom_line(color = "steelblue", size = 1) +
#   theme_minimal() +
#   labs(title = "Collection Effort Over Time",
#        subtitle = "Peak in 1980s–2000s; recent decline likely due to lag in data processing",
#        x = "Year", y = "Number of Records") +
#   theme(plot.title = element_text(face = "bold", size = 14)) +
#   scale_x_continuous(breaks = seq(1700, 2026, 50))
# 
# ggsave("figures/07_timeline.png", p_timeline, width = 14, height = 7, dpi = 300)
# cat("✓ Saved: 07_timeline.png\n")
# 
# # --- 8. RECORDS BY DECADE AND SOURCE
# cat("Creating decade × source heatmap...\n")
# 
# decade_source <- db %>%
#   filter(!is.na(year) & !is.na(source_group)) %>%
#   mutate(year = as.numeric(year)) %>%
#   filter(year >= 1692 & year <= 2026) %>%
#   mutate(decade = floor(year / 10) * 10) %>%
#   count(decade, source_group) %>%
#   arrange(decade)
# 
# p_heatmap <- ggplot(decade_source, aes(x = decade, y = source_group, fill = n)) +
#   geom_tile() +
#   scale_fill_viridis_c(direction = -1) +
#   theme_minimal() +
#   labs(title = "Collection Effort: Decade × Source Group",
#        x = "Decade", y = "Source Group", fill = "Records") +
#   theme(plot.title = element_text(face = "bold", size = 14),
#         axis.text.x = element_text(angle = 45, hjust = 1))
# 
# ggsave("figures/08_decade_source_heatmap.png", p_heatmap, width = 12, height = 7, dpi = 300)
# cat("✓ Saved: 08_decade_source_heatmap.png\n")
# 
# # ============================================================================
# # DATA QUALITY VISUALIZATIONS
# # ============================================================================
# 
# cat("\n=== CREATING DATA QUALITY VISUALIZATIONS ===\n")
# 
# # --- 9. FIELD COMPLETENESS
# cat("Creating field completeness chart...\n")
# 
# key_fields <- c("decimalLatitude", "decimalLongitude", "year", "locality", 
#                 "minimumElevationInMeters", "family", "genus")
# 
# completeness <- tibble(
#   Field = key_fields,
#   Completeness = sapply(key_fields, function(col) {
#     sum(!is.na(db[[col]])) / nrow(db) * 100
#   })
# ) %>%
#   arrange(Completeness)
# 
# p_completeness <- ggplot(completeness, aes(x = reorder(Field, Completeness), y = Completeness, fill = Completeness)) +
#   geom_col() +
#   coord_flip() +
#   scale_fill_viridis_c(direction = -1) +
#   geom_text(aes(label = paste0(round(Completeness, 1), "%")), hjust = -0.1, size = 4) +
#   theme_minimal() +
#   ylim(0, 110) +
#   labs(title = "Field Completeness Across Key Data Fields",
#        subtitle = "Strong geographic core (coords, year), weaker periphery (elevation, genus)",
#        x = "Field", y = "Completeness (%)", fill = "Completeness") +
#   theme(plot.title = element_text(face = "bold", size = 14),
#         legend.position = "bottom")
# 
# ggsave("figures/09_field_completeness.png", p_completeness, width = 11, height = 7, dpi = 300)
# cat("✓ Saved: 09_field_completeness.png\n")
# 
# # --- 10. DATA QUALITY BY SOURCE GROUP
# cat("Creating quality by source group...\n")
# 
# quality_by_source <- db %>%
#   group_by(source_group) %>%
#   summarise(
#     coords = sum(!is.na(decimalLatitude) & !is.na(decimalLongitude)) / n() * 100,
#     year = sum(!is.na(year)) / n() * 100,
#     locality = sum(!is.na(locality)) / n() * 100,
#     elevation = sum(!is.na(minimumElevationInMeters)) / n() * 100,
#     .groups = "drop"
#   ) %>%
#   pivot_longer(-source_group, names_to = "field", values_to = "completeness")
# 
# p_quality_source <- ggplot(quality_by_source, aes(x = source_group, y = completeness, fill = field)) +
#   geom_col(position = "dodge") +
#   coord_flip() +
#   scale_fill_viridis_d(option = "turbo") +
#   theme_minimal() +
#   labs(title = "Data Quality by Source Group",
#        x = "Source Group", y = "Field Completeness (%)", fill = "Field") +
#   theme(plot.title = element_text(face = "bold", size = 14))
# 
# ggsave("figures/10_quality_by_source.png", p_quality_source, width = 12, height = 8, dpi = 300)
# cat("✓ Saved: 10_quality_by_source.png\n")
# 
# # ============================================================================
# # SOURCE COMPOSITION
# # ============================================================================
# 
# cat("\n=== CREATING SOURCE COMPOSITION VISUALIZATIONS ===\n")
# 
# # --- 11. SOURCE GROUP COMPOSITION (waffle-style or donut)
# cat("Creating source composition chart...\n")
# 
# source_comp <- db %>%
#   count(source_group) %>%
#   arrange(desc(n)) %>%
#   mutate(pct = round(100 * n / sum(n), 1))
# 
# p_source <- ggplot(source_comp, aes(x = "", y = n, fill = source_group)) +
#   geom_col(width = 1) +
#   coord_polar("y", start = 0) +
#   geom_text(aes(label = paste0(source_group, "\n", pct, "%")), 
#             position = position_stack(vjust = 0.5), size = 4) +
#   scale_fill_viridis_d(option = "turbo") +
#   theme_void() +
#   labs(title = "Data Source Composition",
#        subtitle = sprintf("Total: %d records across %d sources", nrow(db), nrow(source_comp))) +
#   theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
#         plot.subtitle = element_text(hjust = 0.5),
#         legend.position = "none")
# 
# ggsave("figures/11_source_composition.png", p_source, width = 10, height = 8, dpi = 300)
# cat("✓ Saved: 11_source_composition.png\n")
# 
# # --- 12. SOURCE GROUP BAR CHART
# cat("Creating source group bar chart...\n")
# 
# p_source_bar <- ggplot(source_comp, aes(x = reorder(source_group, n), y = n, fill = n)) +
#   geom_col() +
#   coord_flip() +
#   scale_fill_viridis_c(direction = -1) +
#   geom_text(aes(label = paste0(n, " (", pct, "%)")), hjust = -0.1, size = 4) +
#   theme_minimal() +
#   ylim(0, max(source_comp$n) * 1.1) +
#   labs(title = "Records by Source Group",
#        x = "Source Group", y = "Number of Records", fill = "Records") +
#   theme(plot.title = element_text(face = "bold", size = 14))
# 
# ggsave("figures/12_source_group_distribution.png", p_source_bar, width = 12, height = 8, dpi = 300)
# cat("✓ Saved: 12_source_group_distribution.png\n")
# 
# # ============================================================================
# # SUMMARY STATISTICS TABLE
# # ============================================================================
# 
# cat("\n=== CREATING SUMMARY TABLE ===\n")
# 
# summary_stats <- tibble(
#   Metric = c(
#     "Total Records",
#     "Records with Valid Coordinates",
#     "Unique Families",
#     "Unique Genera",
#     "Unique Species",
#     "Countries Represented",
#     "Year Range",
#     "Median Year",
#     "Source Groups"
#   ),
#   Value = c(
#     format(nrow(db), big.mark = ","),
#     format(nrow(db_coords), big.mark = ","),
#     n_distinct(db$family, na.rm = TRUE),
#     n_distinct(db$genus, na.rm = TRUE),
#     n_distinct(db$scientificName, na.rm = TRUE),
#     n_distinct(db$country, na.rm = TRUE),
#     paste0(min(db$year, na.rm = TRUE), " – ", max(db$year, na.rm = TRUE)),
#     round(median(as.numeric(db$year), na.rm = TRUE)),
#     n_distinct(db$source_group, na.rm = TRUE)
#   )
# )
# 
# write_csv(summary_stats, "figures/summary_statistics.csv")
# cat("✓ Saved: summary_statistics.csv\n")
# 
# cat("\n" %+% paste(rep("=", 80), collapse = ""))
# cat("\n✓ ALL VISUALIZATIONS CREATED SUCCESSFULLY")
# cat("\nOutput folder: figures/\n")
# cat("\nGenerated files:\n")
# system("ls -1 figures/ | grep -E '\\.(png|csv)$'")

```




