---
title: "Data Exploration – EpIG-DB Occurrences"
author: "Francisco Riaño"
date: "today"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: "cosmo"
execute:
  warning: false
  message: false
---

## Data Summary

The present section summarizes the main statistical attributes of the **EpIG-DB** v2.0 occurrences latest version.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(knitr)
library(kableExtra)
library(scales)
library(here)
library(rnaturalearth)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Load EpIG 2.0 complete database (1,372,136 records — full product)
dbPath <- "C:/EpIG_v2.0_/EpIG_volunteership/data/Occurrences_visualization/db_EpIG2_clean_all.rds"
db <- readRDS(here(dbPath))

```

### 1. Dataset Overview

```{r}
#| label: tbl-summary-stats
#| tbl-cap: Summary Statistics of Occurrence Records
total_records <- nrow(db)
valid_coords <- sum(!is.na(db$decimalLatitude) & !is.na(db$decimalLongitude))
pct_valid <- round(valid_coords / total_records * 100, 1)

summary_stats <- tibble(
  Metric = c(
    "Total Records",
    "Records with Valid Coordinates (%)",
    "Unique Species",
    "Unique Genera",
    "Unique Families",
    "Countries Represented",
    "Year Range",
    "Median Collection Year", 
    "Unique Institutions", 
    "Unique Collectors", 
    "Records with Elevation"
  ),
  Value = c(
    format(nrow(db), big.mark = ","),
    paste0(pct_valid, "%"),
    n_distinct(db$scientificName, na.rm = TRUE),
    n_distinct(db$genus, na.rm = TRUE),
    n_distinct(db$family, na.rm = TRUE),
    n_distinct(db$country, na.rm = TRUE),
    paste0(min(db$year, na.rm = TRUE), " – ", max(db$year, na.rm = TRUE)),
    round(median(as.numeric(db$year), na.rm = TRUE)), 
    n_distinct(db$institutionCode, na.rm = TRUE), 
    n_distinct(db$recordedBy, na.rm = TRUE), 
    paste0(
    format(sum(!is.na(db$minimumElevationInMeters)), big.mark = ","),
    " (", 
    round(sum(!is.na(db$minimumElevationInMeters)) / nrow(db) * 100, 1),
    "%)"
    )
  )
)

# Same table with kableExtra enhancements
summary_stats %>%
  kable(
    #caption = "Table 1. Summary statistics of occurrence records",
    align = c("l", "r"),  # Left-align metric names, right-align values
    col.names = c("Metric", "Value")  # Explicit column names
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "left"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f0f0f0") %>%  # Bold header with gray background
  row_spec(7, italic = TRUE) %>%  # Italicize year range for visual distinction
  column_spec(2, bold = FALSE, width = "15em")  # Set width for value column
```

### 2. Data Completeness

```{r}
#| label: tbl-field-completeness
#| tbl-cap: Field Completeness of Occurrence
key_fields <- c("decimalLatitude", "decimalLongitude", "year", "locality", 
                "minimumElevationInMeters", "family", "genus", "scientificName")

# Add human-readable labels
field_labels <- c(
  "decimalLatitude" = "Latitude",
  "decimalLongitude" = "Longitude",
  "year" = "Year",
  "locality" = "Locality",
  "minimumElevationInMeters" = "Elevation",
  "family" = "Family",
  "genus" = "Genus",
  "scientificName" = "Scientific Name"
)

completeness <- tibble(
  Field = key_fields,
  Count = sapply(key_fields, function(col) sum(!is.na(db[[col]]))),
  Completeness_Percent = sapply(key_fields, function(col) {
    round(sum(!is.na(db[[col]])) / nrow(db) * 100, 1)
  })
) %>%
  mutate(Field_Label = field_labels[Field]) %>%
  arrange(desc(Completeness_Percent)) %>%
  select(Field_Label, Count, Completeness_Percent)

completeness %>%
  kable(
    #caption = "Table 2. Field completeness of occurrence records",
    align = c("l", "r", "r"),
    col.names = c("Field", "Records Present", "Completeness (%)")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "left"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f0f0f0") %>%
  # Color-code percentages
  column_spec(3, 
    color = case_when(
      completeness$Completeness_Percent >= 90 ~ "#2e7d32",  # Green
      completeness$Completeness_Percent >= 70 ~ "#ed6c02",  # Orange
      completeness$Completeness_Percent >= 50 ~ "#ed6c02",  # Orange
      TRUE ~ "#d32f2f"  # Red
    ),
    bold = TRUE
  ) %>%
  # Add footer note
  footnote(
    general = "Fields are ordered from most to least complete.",
    general_title = "Note: ",
    footnote_as_chunk = TRUE
  )
```
### 3. Data Quality Indicators
```{r}
### Data quality indicators
#| label: tbl-quality-flags
#| tbl-cap: Data Quality Flags
quality_flags <- tibble(
  Issue = c(
    "Records with missing coordinates",
    "Records with missing year",
    "Records with missing country",
    "Records with missing scientific name"
  ),
  Count = c(
    sum(is.na(db$decimalLatitude) | is.na(db$decimalLongitude)),
    sum(is.na(db$year)),
    sum(is.na(db$country)),
    sum(is.na(db$scientificName))
  ),
  Percentage = round(Count / nrow(db) * 100, 2)
)

quality_flags %>%
  kable(
    caption = "Table 3. Data quality flags",
    align = c("l", "r", "r"),
    col.names = c("Data Quality Issue", "Records Affected", "Percentage (%)")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "left"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f0f0f0") %>%
  column_spec(3,
    color = case_when(
      quality_flags$Percentage <= 5 ~ "#2e7d32",
      quality_flags$Percentage <= 20 ~ "#ed6c02",
      TRUE ~ "#d32f2f"
    ),
    bold = TRUE
  )
```
#### 3.1. Data Source

This table reveals which data-sets contribute the most to the data-base and their characteristics
```{r}
### Data-set contribution
#| label: tbl-data-source
#| tbl-cap: Contribution by Dataset 
#| 
dataset_summary <- db %>% 
  group_by(datasetID) %>% 
  summarise( 
    Records = n(), 
    Species = n_distinct(scientificName, na.rm = TRUE), 
    Valid_Coords = sum(!is.na(decimalLatitude) & !is.na(decimalLongitude)), 
    Completeness = round(Valid_Coords/Records*100, 1)) %>% 
  arrange(desc(Records)) %>%  
  mutate( 
    Records = format(Records, big.mark = ","), 
    Valid_Coords = format(Valid_Coords, big.mark = ","))

dataset_summary %>%
  kable(
    caption = "Table 4. Contribution by dataset",
    align = c("l", "r", "r", "r", "r"),
    col.names = c("Dataset ID", "Records", "Species", "Valid Coordinates", "Completeness (%)")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "left"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f0f0f0")


```
#### 3.2 Taxonomic Coverage

This table shows what groups are well-represented. 

```{r}
family_summary <- db %>% 
  filter(!is.na(family)) %>% 
  group_by(family) %>% 
  summarise( 
    Records = n(), 
    Genera = n_distinct(genus, na.rm = TRUE), 
    Species = n_distinct(scientificName, na.rm = TRUE)
    ) %>% 
  arrange(desc(Records)) %>%  
  head(10) %>%  
  mutate(Records = format(Records, mig.mark = ","))

family_summary %>%
  kable(
    caption = "Table 5. Top 10 families by record count",
    align = c("l", "r", "r", "r"),
    col.names = c("Family", "Records", "Genera", "Species")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "left"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f0f0f0")
```
### 4. Temporal Distribution

This table summarizes the data collection efforts over time.
```{r}
### Temporal distribution summary
# Decade aggregation
### Temporal distribution summary

# Extract year from eventDate
year_data <- db %>%
  mutate(
    year_extracted = as.numeric(substr(as.character(eventDate), 1, 4))
  ) %>%
  filter(!is.na(year_extracted))

year_summary <- tibble(
  Metric = c(
    "Year Range",
    "Total Years with Records",
    "Number of Records with Year",
    "Records with Missing Year"
  ),
  Value = c(
    paste0(min(year_data$year_extracted, na.rm = TRUE), " – ", 
           max(year_data$year_extracted, na.rm = TRUE)),
    n_distinct(year_data$year_extracted, na.rm = TRUE),
    format(nrow(year_data), big.mark = ","),
    format(nrow(db) - nrow(year_data), big.mark = ",")
  )
)

year_summary %>%
  kable(
    caption = "Table 6. Temporal distribution summary",
    align = c("l", "r"),
    col.names = c("Metric", "Value")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "left"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f0f0f0")
```
#### 4.1 Records Collected by Decade

```{r}
### Temporal distribution summary

# Extract year from eventDate and create decade groupings
temporal_summary <- db %>%
  mutate(
    # Extract year from eventDate (first 4 characters)
    year_extracted = as.numeric(substr(as.character(eventDate), 1, 4)),
    # Create decade from extracted year
    Decade = floor(year_extracted / 10) * 10,
    # Create decade label as range
    Decade_Label = paste0(Decade, " - ", Decade + 10)
  ) %>%
  # Filter out records with missing or invalid years
  filter(!is.na(Decade) & !is.na(year_extracted)) %>%
  group_by(Decade_Label, Decade) %>%
  summarise(
    Records = n(),
    Species = n_distinct(scientificName, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(Decade) %>%
  mutate(Records = format(Records, big.mark = ",")) %>%
  select(-Decade)  # Remove helper column after sorting

# Display the table
temporal_summary %>%
  kable(
    caption = "Table 6. Records by decade",
    align = c("l", "r", "r"),
    col.names = c("Decade", "Records", "Unique Species")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "left"
  ) %>%
  row_spec(0, bold = TRUE, background = "#f0f0f0")
```