---
title: "Occurrence Maps"
author: "Francisco Riaño"
date: "today"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: "cosmo"
execute:
  warning: false
  message: false
---

## Proposed Design

-   Prepare one reusable set of coordinate-valid records within the Neotropical boundary.
-   Display geographic distribution as rasterized point panels separated by source group.
-   Aggregate records into an equal-area hexagonal grid rather than plot 1.37 million overlapping points.
-   Use the dissolved Neotropical boundary as the mask and outline.
-   Use a logarithmic color scale for record counts.
-   Retain empty areas so sampling gaps remain visible.
-   Report the number of coordinate-valid records used and records outside the Neotropical boundary.
-   Avoid exposing individual occurrence coordinates in the rendered output.

## Loading Data

The complete EpIG-DB 2.0 occurrence product is loaded locally. The raw database is employed only to prepare aggregate outputs and is not displayed in the HTML document.

```{r}
#| label: setup
#| echo: false

library(tidyverse)
library(sf)
library(knitr)
library(kableExtra)
library(scales)
library(here)
library(rnaturalearth)
library(dplyr)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Load EpIG 2.0 complete database (1,372,136 records — full product)
dbPath <- "C:/EpIG_v2.0_/EpIG_volunteership/data/Occurrences_visualization/db_EpIG2_clean_all.rds"
db <- readRDS(here(dbPath))
```

## Spatial Data Preparation

### Spatial Layers

Three spatial inputs are prepared. the dissolved Neotropical boundary, country boundaries, and the Morrone biogeographic provinces. Invalid geometries are repaired before further spatial operations.

```{r}
#| echo: false
#| label: prepare-map-layers

library(sf)
library(dplyr)

map_dir <- file.path(
  "C:/EpIG_v2.0_",
  "EpIG_volunteership",
  "data",
  "Occurrences_visualization",
  "Neotropics_Morrone"
)

# Read source layers
# -----------------------------------------------
neotropics_raw <- sf::st_read(
  file.path(map_dir, "Neotropic_dissolved.shp"),
  quiet = TRUE
)
# Country boundaries
countries_raw <- sf::st_read(
  file.path(map_dir, "Neotropic_Paises.shp"),
  quiet = TRUE
)
# Morrone biogeographic provinces
provinces_raw <- sf::st_read(
  file.path(map_dir, "NeotropicMap_WGS84_1.shp"),
  quiet = TRUE
)

# Record source-layer properties before geometry repair
map_layer_summary <- tibble(
  Layer = c(
    "Neotropical boundary",
    "Countries",
    "Biogeographic provinces"
  ),
  Purpose = c(
    "Analysis mask and outer boundary",
    "Political reference boundaries",
    "Morrone biogeographic units for subsequent maps"
  ),
  Features = c(
    nrow(neotropics_raw),
    nrow(countries_raw),
    nrow(provinces_raw)
  ),
  `Invalid geometries before repair` = c(
    sum(!sf::st_is_valid(neotropics_raw)),
    sum(!sf::st_is_valid(countries_raw)),
    sum(!sf::st_is_valid(provinces_raw))
  ),
  EPSG = c(
    sf::st_crs(neotropics_raw)$epsg,
    sf::st_crs(countries_raw)$epsg,
    sf::st_crs(provinces_raw)$epsg
  )
)

# Repair geometries 
neotropics_raw <- sf::st_make_valid(neotropics_raw)
countries <- sf::st_make_valid(countries_raw)
provinces <- sf::st_make_valid(provinces_raw)

# Dissolve the Neotropical boundary into one geometry
neotropics <- sf::st_sf(
  geometry = sf::st_union(sf::st_geometry(neotropics_raw)),
  crs = sf::st_crs(neotropics_raw)
)
```

Table @tbl-map-layers documents the role, size, geometry status, and coordinate reference system of each spatial input. All three inputs use WGS84 (EPSG:4326).

```{r}
#| label: tbl-map-layers
#| echo: false
#| tbl-cap: "Spatial layers used in the occurrence-map workflow."

map_layer_summary |>
  knitr::kable(
    format = "html",
    align = c("l", "l", "r", "r", "r"),
    format.args = list(big.mark = ",")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )
```

### Coordinate Screening

Longitude and latitude are converted to numeric values and evaluated sequentially. Records must have a complete coordinate pair, fall within the legal geographic ranges, and not represent the placeholder coordinate '(0, 0)'.

```{r}
#| label: prepare-occurrence-points
#| echo: false

# Fields that potentially will be required by the map workflow
map_candidate_fields <- c(
  "occurrenceID",
  "datasetID",
  "source_group",
  "family",
  "genus",
  "scientificName",
  "matchedNameEpig",
  "taxonRank",
  "eventDate",
  "decimalLongitude",
  "decimalLatitude"
)

# Confirm which candidate fields occur in the complete database
map_field_inventory <- tibble(
  Field = map_candidate_fields,
  Available = map_candidate_fields %in% names(db)
)

# Coordinates employed for every map
required_coordinate_fields <- c(
  "decimalLongitude",
  "decimalLatitude"
)

missing_coordinate_fields <- setdiff(
  required_coordinate_fields,
  names(db)
)

if (length(missing_coordinate_fields) > 0) {
  stop(
    paste(
      "Missing required coordinate fields:",
      paste(missing_coordinate_fields, collapse = ", ")
    )
  )
}

# Retain only fields needed for the map workflow
coordinate_audit <- db |>
  mutate(
    record_index = dplyr::row_number(),
    .before = 1
  ) |>
  select(
    record_index,
    any_of(map_candidate_fields)
  ) |>
  mutate(
    longitude = suppressWarnings(
      as.numeric(as.character(decimalLongitude))
    ),
    latitude = suppressWarnings(
      as.numeric(as.character(decimalLatitude))
    ),
    has_coordinate_pair =
      !is.na(longitude) &
      !is.na(latitude),
    within_legal_range =
      has_coordinate_pair &
      is.finite(longitude) &
      is.finite(latitude) &
      between(longitude, -180, 180) &
      between(latitude, -90, 90),
    is_zero_zero =
      within_legal_range &
      longitude == 0 &
      latitude == 0,
    is_coordinate_valid =
      within_legal_range &
      !is_zero_zero
  ) |>
  select(
    -decimalLongitude,
    -decimalLatitude
  )

total_records <- nrow(coordinate_audit)

complete_pairs <- sum(
  coordinate_audit$has_coordinate_pair
)

outside_legal_range <- sum(
  coordinate_audit$has_coordinate_pair &
    !coordinate_audit$within_legal_range
)

zero_zero_records <- sum(
  coordinate_audit$is_zero_zero
)

valid_coordinate_records <- sum(
  coordinate_audit$is_coordinate_valid
)

coordinate_screening_summary <- tibble(
  Screening_step = c(
    "All occurrence records",
    "Excluded: incomplete or non-numeric coordinate pair",
    "Excluded: coordinate outside legal range",
    "Excluded: placeholder coordinate (0, 0)",
    "Retained: coordinate-valid records"
  ),
  Records = c(
    total_records,
    total_records - complete_pairs,
    outside_legal_range,
    zero_zero_records,
    valid_coordinate_records
  )
) |>
  mutate(
    `Percentage of all records` = scales::percent(
      Records / total_records,
      accuracy = 0.1
    )
  )

# Convert the coordinate-valid subset to an sf object while
# retaining the analytical attributes
occurrence_geo <- coordinate_audit |>
  filter(is_coordinate_valid) |>
  select(
    record_index,
    any_of(
      c(
        "occurrenceID",
        "datasetID",
        "source_group",
        "family",
        "genus",
        "scientificName",
        "matchedNameEpig",
        "taxonRank",
        "eventDate"
      )
    ),
    longitude,
    latitude
  ) |>
  sf::st_as_sf(
    coords = c("longitude", "latitude"),
    crs = 4326,
    remove = FALSE
  )
```

Table @tbl-coordinate-screening shows how coordinate-valid subset was obtained. Exclusion categories are mutually exclusive within this sequence.

```{r}
#| label: tbl-coordinate-screening
#| echo: false
#| tbl-cap: "Sequential screening of occurrence coordinates."

coordinate_screening_summary |>
  knitr::kable(
    format = "html",
    align = c("l", "r", "r"),
    format.args = list(big.mark = ","),
    col.names = c(
      "Screening step",
      "Records",
      "Percentage of all records"
    )
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )
```

### Projected Neotropical Dataset

Coordinate-valid records are projected to Equal Earth (EPSG:8857), filtered by the dissolved Neotropical boundary, and assigned to a 100-km equal-area hexagonal grid. Employing an equal-are projection makes the grid cells spatially comparable across the study region.

```{r}
#| label: prepare-spatial-subset
#| echo: false


# Equal-area projection suitable for cell-based comparisons
map_crs_equal_area <- 8857

neotropics_equal <- st_transform(
  neotropics,
  crs = map_crs_equal_area
)

countries_equal <- st_transform(
  countries,
  crs = map_crs_equal_area
)

provinces_equal <- st_transform(
  provinces,
  crs = map_crs_equal_area
)

# Keep all the needed information for posterior maps visualizations
occurrence_points_equal <- occurrence_geo |>
  sf::st_transform(crs = map_crs_equal_area)

# Retain records inside the Neotropical boundary
inside_neotropics <- lengths(
  st_intersects(occurrence_points_equal, neotropics_equal)
) > 0

occurrence_neotropics <- occurrence_points_equal[
  inside_neotropics,
]

```
## Geographic Distribution

### Distribution by Data Source

The geographic distribution of coordinate-valid records was compared among four broad source groups. Separate panels were used because repositories account for most recods and would obscure less sources in a single overlaid map. All panels use the same projection, geographic extent, point size, and transparency. 

```{r}
#| label: prepare-geographic-distribution
#| echo: false

# Prepare a regular data frame for the geographic-distribution figure.
#
# occurrence_neotropics is an sf object:
# - each row represents one retained occurrence record;
# - the geometry column contains its projected point location;
# - the remaining columns contain attributes such as source_group.

# st_geometry() extracts only the spatial geometry from the sf object.
# st_coordinates() converts those point geometries into a numeric matrix:
#   column 1 = x coordinate
#   column 2 = y coordinate

# 1. Extract the projected Coordinates --------
# Extract Equal Earth coordinates from the retained point geometries
point_coordinates <- sf::st_coordinates(
  sf::st_geometry(occurrence_neotropics)
)

# 1. Build the Plotting DataFrame ---------
# st_drop_geometry() removes the sf geometry column while retaining the
# ordinary attribute columns.
#
# transmute() creates a new data frame containing only the variables
# declared below. This keeps the plotting object small.
geographic_distribution_data <- occurrence_neotropics |>
  sf::st_drop_geometry() |>
  transmute(
    record_index,
    source_group = stringr::str_to_lower(
      stringr::str_trim(source_group)
    ),
    x = point_coordinates[, 1],
    y = point_coordinates[, 2]
  )

# 3. Define the expected source groups and their display order
source_group_levels <- c(
  "repositories",
  "researchers",
  "herbaria",
  "literature"
)

# 4.Check for unexpected values --------
unexpected_source_groups <- setdiff(
  unique(geographic_distribution_data$source_group),
  source_group_levels
)
# Stop the render if an unexpected category is detected
if (length(unexpected_source_groups) > 0) {
  stop(
    paste(
      "Unexpected source-group values:",
      paste(unexpected_source_groups, collapse = ", ")
    )
  )
}
# 5. Count Records in each Source Group ------
source_group_counts <- geographic_distribution_data |>
  count(
    source_group,
    name = "records"
  ) |>
  mutate(
    source_group = factor(
      source_group,
      levels = source_group_levels
    ),
    # Construct the text displayed above each map panel
    panel_label = paste0(
      stringr::str_to_title(as.character(source_group)),
      "\n",
      scales::comma(records),
      " records"
    )
  ) |>
  arrange(source_group)
# 6. Attach the Panel Labels to Every Record --------
geographic_distribution_data <- geographic_distribution_data |>
  left_join(
    source_group_counts |>
      select(source_group, panel_label),
    by = "source_group"
  ) |>
  mutate(
    source_group = factor(
      source_group,
      levels = source_group_levels
    ),
    panel_label = factor(
      panel_label,
      levels = source_group_counts$panel_label
    )
  )
# 7. Define the Map colors ---------
source_group_palette <- c(
  repositories = "#0072B2",
  researchers = "#D55E00",
  herbaria = "#009E73",
  literature = "#CC79A7"
)
```

```{r}
#| label: fig-geographic-distribution
#| echo: false
#| fig-cap: "Geographic distribution of coordinate-valid EpIG-DB occurrence records within the Neotropical region, separated by source group."
#| fig-width: 11
#| fig-height: 9
#| fig-format: png
#| dpi: 300

ggplot() +
  geom_sf(
    data = neotropics_equal,
    fill = "grey98",
    colour = NA
  ) +
  geom_point(
    data = geographic_distribution_data,
    aes(
      x = x,
      y = y,
      colour = source_group
    ),
    size = 0.08,
    alpha = 0.15,
    shape = 16
  ) +
  geom_sf(
    data = countries_equal,
    fill = NA,
    colour = "grey45",
    linewidth = 0.18
  ) +
  geom_sf(
    data = neotropics_equal,
    fill = NA,
    colour = "black",
    linewidth = 0.4
  ) +
  facet_wrap(
    vars(panel_label),
    ncol = 2
  ) +
  scale_colour_manual(
    values = source_group_palette,
    guide = "none"
  ) +
  coord_sf(
    crs = sf::st_crs(neotropics_equal),
    datum = NA,
    expand = FALSE
  ) +
  labs(
    title = "Geographic distribution of EpIG-DB records",
    subtitle = "Coordinate-valid records by source group",
    caption = paste0(
      "Points represent occurrence records rather than unique localities. ",
      "Overlapping records may appear as areas of greater visual intensity."
    )
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title = element_text(
      face = "bold",
      size = 14
    ),
    plot.subtitle = element_text(
      size = 11
    ),
    plot.caption = element_text(
      size = 8,
      colour = "grey35",
      hjust = 0
    ),
    strip.text = element_text(
      face = "bold",
      size = 10,
      margin = margin(4, 4, 4, 4)
    ),
       panel.spacing = unit(
      0.8,
      "lines"
    )
  )
```

## Sampling Density

### Spatial Aggregation

Records retained within tyhe Neotropical boundary are assigned to a 100-km hexagonal grid in the Equal Earth Projection. Interior cells are spatially comparable; cells intersecting the outer biundary are clipped for display. 

```{r}
#| label: calculate-occurrence-density
#| echo: false

# Construct a 100-km hexagonal grid
grid_geometry <- st_make_grid(
  neotropics_equal,
  cellsize = 100000,
  square = FALSE
)

occurrence_grid <- st_sf(
  grid_id = seq_along(grid_geometry),
  geometry = grid_geometry
)

# Retain grid cells touching the Neotropical boundary
grid_inside <- lengths(
  st_intersects(occurrence_grid, neotropics_equal)
) > 0

occurrence_grid <- occurrence_grid[grid_inside, ]

# Assign each occurrence to a grid cell
point_cell_matches <- st_intersects(
  occurrence_neotropics,
  occurrence_grid
)

point_cell_id <- vapply(
  point_cell_matches,
  function(x) {
    if (length(x) == 0) NA_integer_ else x[[1]]
  },
  integer(1)
)

# Count occurrence records per cell
cell_counts <- tabulate(
  point_cell_id[!is.na(point_cell_id)],
  nbins = nrow(occurrence_grid)
)

occurrence_grid$n_records <- cell_counts

# Clip all cells, including empty cells, to the study boundary
occurrence_grid_plot <- suppressWarnings(
  sf::st_intersection(
    occurrence_grid,
    neotropics_equal
  )
) |>
  mutate(
    n_records_plot = na_if(n_records, 0L)
  )

records_inside <- nrow(occurrence_neotropics)
records_assigned <- sum(occurrence_grid$n_records)

# Simple quality-control summary
density_map_summary <- tibble(
  Metric = c(
    "Coordinate-valid records",
    "Records inside the Neotropical boundary",
    "Records outside the Neotropical boundary",
    "Records assigned to grid cells",
    "Occupied grid cells",
    "Empty grid cells",
    "Cell-assignment check"
  ),
  Value = c(
    scales::comma(nrow(occurrence_geo)),
    scales::comma(records_inside),
    scales::comma(nrow(occurrence_geo) - records_inside),
    scales::comma(records_assigned),
    scales::comma(sum(occurrence_grid$n_records > 0)),
    scales::comma(sum(occurrence_grid$n_records == 0)),
    if_else(
      records_inside == records_assigned,
      "Passed",
      "Review required"
    )
  )
)
```

Table @tbl-density-summary reports the final map denominator and verifies that all records inside the study boundary were assigned to grid cells.

```{r}
#| label: tbl-density-summary
#| echo: false
#| tbl-cap: "Summary of records and grid cells used in the occurrence-density map."

density_map_summary |>
  knitr::kable(
    format = "html",
    align = c("l", "r")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  )
```

### Sampling Density <ap

Figure @fig-occurrence-density displays the number of records in each occupied cell. The logarithmic color transformation preserves differences across a highly differentiated count distribution. Light-grey cells represent sampled-area gaps with no retained occurrence records. 

```{r}
#| label: fig-occurrence-density
#| echo: false
#| fig-cap: "Spatial distribution of EpIG-DB occurrence records across the Neotropical region. Counts were aggregated within 100-km equal-area hexagonal cells: light-grey cells contain no retained records."
#| fig-width: 9
#| fig-height: 8

ggplot() +
  geom_sf(
    data = occurrence_grid_plot,
    aes(fill = n_records_plot),
    colour = "white",
    linewidth = 0.03
  ) +
  geom_sf(
    data = countries_equal,
    fill = NA,
    colour = "grey45",
    linewidth = 0.2
  ) +
  geom_sf(
    data = neotropics_equal,
    fill = NA,
    colour = "black",
    linewidth = 0.45
  ) +
  scale_fill_viridis_c(
    name = "Occurrence\nrecords",
    option = "magma",
    direction = -1,
    trans = "log10",
    labels = scales::label_number(big.mark = ","),
    na.value = "grey92"
  ) +
  coord_sf(datum = NA) +
  labs(
    title = "Geographic coverage of EpIG-DB",
    subtitle = "Occurrence records per 100-km equal-area hexagonal cell",
    caption = paste0(
      "Coordinate-valid records within the dissolved Neotropical boundary. ",
      "Light-grey cells contain no retained records."
    )
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    plot.caption = element_text(
      size = 8,
      colour = "grey35",
      hjust = 0
    ),
    legend.position = "right"
  )
```

### Interpretation

The map summarizes geographic sampling intensity rather than species richness or biological abundance. Cells with many records may reflect greater collection effort, data mobilization, or repeated sampling. Consequently, geographic differences in record density should not be interpreted directly as differences in epiphyte diversity. 
